Efficient Data Provenance Analysis for Concurrent Programs

### 2013-11-12 paper.tex
committing many changes from a read on 11/11.  Now, prepping to rework related & be sure the eval isn't lying or bad.

\title{Data Provenance for Concurrent Program Debugging and Analysis in Production Systems}

@@

These efforts have shown some of the promise of provenance, but are not a
debugging panacea and are insufficient for dealing with concurrent program
bugs. First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
% Limiting the applicability of their technique to only some failures, like {\tt
% Null Pointer Exceptions}.
Second, while prior work has shown that limited provenance
tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive. Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples}{\bf TODO: Brandon, check this reference, since you just cited it in the previous sentence.} impose overheads too high for production for
C and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems. Addressing
concurrency is important because a value could have originated in any of a
large number of threads, making reasoning about why a variable contains that
value complex and difficult. In summary, the challenges in using provenance to debug concurrent programs are: (1) collecting the right information to characterize the concurrent execution behavior; and (2) making it efficient enough for general programs running in production settings.

@@

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call to an \lwt update
function in our runtime library just before each write that may access data
last written by another thread (escaping memory accesses).  Write
instrumentation calls are passed three pieces of information: the address of
the memory word involved in the access, the thread identifier of the accessing
thread, and the accessing program point.

We use a {\tt GCC}'s local function escape analysis to identify accesses to
memory locations that are provably inaccessible to code running in other
threads.  Such operations access unshared data and they are not instrumented by
our compiler.  
%Using an inter-procedural thread escape analysis may eliminate more
%instrumentation sites.

\subsection{\lwt Formalism and Correctness}
{\bf TODO: Necessary?  Is it really interesting?}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = \{q_{0}, \ldots, q_{e}\}$ where each $q \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

@@

Our implementation of the \lwt does not add any synchronization to the program.
As a result our analysis does not incur the overheads or possible thread
schedule perturbation synchronization can cause.  However, there are two
potential issues that stem from this design choice. First, we do not enforce
memory consistency of meta-data reads and writes in different threads.  Second,
our implementation does not enforce the atomicity of \lwt updates and their
corresponding program accesses.  In the absence of data-races our \lwt
implementation is {\em always correct} and neither of these issues exist.  Both
atomicity and ordering is enforced by the program synchronization that ensures
consistency for the data being written by the instructions our system
instruments.

We use our \lwt formalism to show that without data-races, \lwt updates and
their corresponding memory accesses are atomic and well-ordered.  The relation
$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
order.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
happens before all accesses that $p^{t}_{j}$ happens before {\em and before
their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
after all preceding accesses to $m$ in other threads, {\em and after their
repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.

@@

%{\bf TODO: Figure?}

@@

\ctraps is a framework for analyzing and interposing on inter-thread
communication events during a program's execution.  We build \ctraps on top of
last writer slicing that we described in Section~\ref{sec:lastwriterslices}.
Using last writer slice information, \ctraps dynamically tracks 
inter-thread communication events. \ctraps exposes communication
events as {\em traps} during an execution that {\em trap handlers} can handle
to perform analysis and interposition.  We formalize trap delivery and handling
in Section~\ref{sec:ctsoundness}.

@@

\ctraps delivers software traps before all read and write operations executed
by one thread that access a memory location that was last written by another
thread, {\em i.e.}, communicating memory operations.  Traps are simply function
calls made before such operations execution.  \ctraps implements traps using a
combination of compiler and runtime library support.  Note that we use the name
``trap'' for the interposition points in \ctraps, but we implement traps using 
function calls only, as opposed to signals or some other mechanism.

\paragraph{Compiler Support}
Our compiler inserts a call to the \ctraps {\em trap hook} function immediately
before all read and write operations that access potentially escaping memory
locations.  Trap hooks are passed the address of the memory location being
accessed, the thread identifier of the accessing thread, the type of operation
being performed, and the program point of the access.

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function,  which has two purposes.
The trap hook function detects communication and delivers traps to \ctraps
applications when communication occurs.  The trap hook function detects
communication by looking up the \lwt entry for the memory location being
accessed.  If the thread identifier in the \lwt entry is different from the
accessing thread, then the trap hook function delivers a trap to the accessing
thread, just before the memory access is allowed to proceed.  To deliver a
trap, the trap hook function calls {\em trap handler routines} that are defined
in \ctraps applications.  Trap handler routines are like signal handlers.  They
are registered with the \ctraps runtime when the program starts.  Trap handlers
may contain arbitrary application-specific code.  Figure~\ref{fig:hookapi}
shows the \ctraps trap handling API.  Trap handlers are passed the four pieces
of information passed to the trap hook, as well as the \lwt's record of the
program point and thread identifier of the last write to the involved memory
location.

\begin{figure}[htb]
\centering
\begin{verbatim}
void trap(AccessType type, 
          ThdId currThd, ProgPt currPt,
          ThdId lastThd, ProgPt lastPt);
\end{verbatim}
\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
\end{figure}

@@

\subsection{Release}
We released our implementation, free and open-sourced, on the
web~\cite{withheldforanonymity} and published it to the GCC plugin repository.
Since its release, we have seen contributions to our codebase from members of
the GCC development community and downloads from interested developers.

\section{Evaluation}
\label{sec:eval}
We evaluated last writer slices and \ctraps along several axes.  We show that
last writer slices provide information that is useful during debugging in
Section~\ref{sec:eval:debugging} using case studies on several buggy programs.
We evaluate our performance overheads and show that last writer slices and
\ctraps are feasible for use in deployed systems for a majority of workloads we
studied.  We then characterize our instrumentation and discuss the sources of
our system's run time overhead.   We show that \ctraps is useful by
implementing two existing applications and analyzing their performance.

@@

We illustrate the debugging benefits of last writer slices by using our system
to debug six real-world programs studied previously in the debugging
literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
Section~\ref{sec:eval:debugging:cases} looks in detail at several cases to 
better illustrate how last writer slices helps with debugging.

In Table~\ref{tab:bugs}, we show the bugs we studied describing the program,
bug type, and bug report identifier where applicable.  For each bug, we
followed the procedure for installation and bug triggering described by the
author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
in this work, including both atomicity and ordering violations.

@@

\begin{table}
\scriptsize
\centering
\begin{tabular}{l|ll|l|ccc|c}
{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    & Yes   & Yes \\
httrack& 3.43.9  &  N/A     & Order  &Yes  &No    & Yes   & No  \\%Used w/o origin - no help
jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
\end{tabular}
\caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
\normalsize
\end{table}

The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
violation bugs, last writer slices are effective and useful.  In all cases, we
found that the last writer slice revealed the provenance of the data involved
in the operation that caused the failure, so last writer slicing was effective.
We found that none of the failures we looked at were trivially debuggable.
This finding is consistent with the difficulty of debugging concurrency bugs --
the core dump does not reveal the interaction between threads that led to a
failure, only the result of that interaction that leads to the failure.

Last writer slices were obviously useful in five of six cases.  In these cases,
the root cause of the bug becomes clear when last writer slices show the
inter-thread data flow led to a variable having a value that caused a failure.
Using that information, we expect it would be straightforward in these cases to
add synchronization to fix these bugs.  Note that without last writer slices,
understanding the responsible inter-thread data flow and fixing the bug would
be difficult.  For the remaining case, {\tt apache2}, last writer slices would
probably help with debugging, but it is less obvious because this bug is a
complex, multi-variable bug.  We provide more detail on {\tt apache2} in Section~\ref{sec:eval:debugging:cases}.

Our evaluation shows that bad value origin tracking~\cite{badapples} is not
likely to be helpful in five out of six cases where last writer slices is
useful.  In one case, a use-after-delete bug in {\tt pbzip2}, bad value origin
tracking helps.  For this bug, when one thread deletes a variable, it makes
that variable unusable and bad value origin tracking would track that value.
When the failure occurs, the programmer would then know why the value was
deleted.  In the other five cases we looked at, the values involved in causing
the failure are usable and would not be tracked by bad value origin tracking.

@@

\paragraph{JDK-StringBuffer}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:debugging} describes the benefit provided by
last writer slices in debugging an ordering violation in the JDK-1.4.

@@

\paragraph{HTTrack}
%{\bf TODO: Deal with this duplication from the copy/paste}
%Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
%plays in debugging, using an example of a real bug from HTTrack, a web caching
%tool.  The figure depicts two threads manipulating a data structure that
%contains a mutex\_lock.  The program has an ordering violation bug: the
%program crashes when Thread 2 tries to acquire the lock before Thread 1 has
%initialized it.  Assuming the run time environment supports it, a core dump
%delivered to the programmer after such a failure reveals very little about how
%to fix the bug.  The lock is uninitizalized, so the core dump may show that it
%holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
%insight into the root cause of the failure, which is that the code permits a
%use of the lock before its initialization.     This revalation is of limited use as it does not rule out
%arbitrary data corruption as the bug's root cause.
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
%\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
%ordering violation in HTTrack.} The provenance information stored in the last
%writer slice for the mutex variable helps understand the bug's root cause.}
%\end{figure}
%
%
%The vital, missing piece to solving this bug is provenance information.
%Debugging would be easier if the core dump included information about the
%thread and code point that last wrote the lock.  The programmer would see that
%no thread had initialized the lock in the failing execution, but that Thread 1
%initialized the lock in non-failing executions.  This information guides the
%programmer to fix the bug by adding synchronization that ensures the correct
%operation ordering.

Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  The program
crashes at point {\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid
mutex.    The root cause of this bug is that Thread 1 fails to initialize the
mutex before Thread 2 uses it.   The core dump is of little debugging use, only
showing that, indeed, the lock is uninitialized.  Even if the programmer
compared the core dump from the failing run to the memory image from a correct
execution -- a helpful debugging strategy -- they would find only that in a
correct execution, the lock contains valid mutex state.

@@

Looking at last writer slices make this root cause clear: in the failing
execution, the last writer slice indicates the mutex is uninitialized; in the
non-failing execution, the last writer slice indicates the mutex was
initialized by Thread 1 at point {\bf A}. While the core dump provides no
information about the connection between point {\bf A} and point {\bf B}, the
last writer slice reveals that the programmer should ensure {\bf B} follows
{\bf A} to prevent the failure.  Note that even augmenting the core dump with
bad value origin tracking~\cite{badapples} does not help with this bug.  In the
failing execution, origin tracking shows that the unusable lock is
uninitialized, but in the non-failing exeuction, origin tracking does not
monitor the {\em usable} lock value.  By contrast, last writer slicing tracks
that crucial information, helping debug this complex concurrency bug.

@@

We triggered the bug by running an insert query during a log rotation and
execution stopped at {\bf C}.  The core dump showed the log was marked closed,
which is the failure state ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
However, the core dump did not reveal {\em why} the log was marked closed.  The
last writer slice at {\bf C} for {\tt log\_type} shows that it was written last
by Thread 1 at {\bf A}, in the log rotation code, which should be atomic.
Thus, the last writer slice shows the programmer the need for log rotation to
be atomic with respect to the insert code to prevent this failure.

>>>>

\title{Efficient Data Provenance Analysis for Concurrent Programs}

@@

These efforts have shown some of the promise of provenance, but are not a
debugging panacea and are insufficient for dealing with concurrent program
bugs. First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
% Limiting the applicability of their technique to only some failures, like {\tt
% Null Pointer Exceptions}.
Second, while prior work has shown that limited provenance tracking for Java
programs is possible with overheads low enough for production~\cite{badapples},
a more general form of provenance tracking for C and C++ programs has been
elusive. Prior approaches that build on slicing~\cite{thinslicing,
tipslicingsurvey, whylineicse} or heavy-weight memory checkers~\cite{badapples}
impose overheads too high for production for C and C++ programs. Third, no
prior technique that we know of has collected provenance information that is
tailored to debugging concurrent programs with overheads that are low enough
for use in production systems. Addressing concurrency is important because a
value could have originated in any of a large number of threads, making
reasoning about why a variable contains that value complex and difficult. In
summary, the challenges in using provenance to debug concurrent programs are:
(1) collecting the right information to characterize the concurrent execution
behavior; and (2) making it efficient enough for general programs running in
production settings.

@@

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call to an \lwt update
function in our runtime library just before each write that may access data
last written by another thread (escaping memory accesses).  Write
instrumentation calls are passed three pieces of information: the address of
the memory word involved in the access, the thread identifier of the accessing
thread, and the accessing program point.   We use a {\tt GCC}'s local function
escape analysis to identify accesses to memory locations that are provably
inaccessible to code running in other threads.  Such operations access unshared
data and they are not instrumented by our compiler.

%Using an inter-procedural thread escape analysis may eliminate more
%instrumentation sites.

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

@@

Our implementation of the \lwt does not add any synchronization to the program.
As a result our analysis does not incur the overheads or possible thread
schedule perturbation synchronization can cause.  However, there are two
potential issues that stem from this design choice. First, we do not enforce
memory consistency of meta-data reads and writes in different threads.  Second,
our implementation does not enforce the atomicity of \lwt updates and their
corresponding program accesses.  In the absence of data-races our \lwt
implementation is {\em always correct} and neither of these issues exist.  Both
atomicity and ordering is enforced by the program synchronization that ensures
consistency for program data.

We use our \lwt formalism to show that without data-races, \lwt updates and
their corresponding memory accesses are atomic and well-ordered.  The relation
$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
order.  $I(p^{t'}_{j'})$ is inserted before $p^{t'}_{j'}$, but after the synchronization that orders $p^{t'}_{j'}$ and $p^{t}_{j}$, so $p^{t}_{j} \prec I(p^{t'}_{j'})$.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
happens before all accesses that $p^{t}_{j}$ happens before {\em and before
their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
after all preceding accesses to $m$ in other threads, {\em and after their
repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.

@@

@@

\ctraps is a framework for analyzing and interposing on inter-thread
communication events during a program's execution.  We build \ctraps on top of
last writer slicing that we described in Section~\ref{sec:lastwriterslices}.
Using last writer slice information, \ctraps dynamically tracks 
inter-thread communication events. \ctraps exposes communication
events as {\em traps} during an execution that {\em trap handlers} can handle
to perform analysis and interposition.

@@

\ctraps delivers software traps before all read and write operations executed
by one thread that access a memory location that was last written by another
thread, {\em i.e.}, communicating memory operations.  Traps are simply inserted
function calls (as opposed to, {\em e.g.}, signals) made before such operations
execution.  \ctraps implements traps using a combination of compiler and
runtime library support.

\paragraph{Compiler Support}
Our compiler inserts a call to the \ctraps~{\em trap hook} function immediately
before all read and write operations that access potentially escaping memory
locations.  Trap hooks are passed the address of the memory location being
accessed, the thread identifier of the accessing thread, the type of operation
being performed, and the program point of the access.

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function,  which has two purposes.
The trap hook function detects communication and delivers traps to \ctraps
applications when communication occurs.  The trap hook function detects
communication by looking up the \lwt entry for the memory location being
accessed.  If the thread identifier in the \lwt entry is different from the
accessing thread, then the trap hook function delivers a trap to the accessing
thread, just before the memory access is allowed to proceed.  To deliver a
trap, the trap hook function calls {\em trap handler routines} that are defined
in \ctraps applications.  Trap handler routines are like signal handlers.  The
\ctraps runtime maintains a list of trap handlers that are registered when the
program starts via a configuration stored in the environment.  Trap handlers
may contain arbitrary application-specific code.  \ctraps defines a trap
handling API.  Trap handlers are passed the four pieces of information passed
to the trap hook, as well as the \lwt's record of the program point and thread
identifier of the last write to the involved memory location.

%\begin{figure}[htb]
%\centering
%\begin{verbatim}
%void trap(AccessType type, 
%          ThdId currThd, ProgPt currPt,
%          ThdId lastThd, ProgPt lastPt);
%\end{verbatim}
%\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
%\end{figure}

@@

\subsection{Release}
We released our implementation, free and open-sourced, on the
web~\cite{ctrapsrelease} and published it to the GCC plugin repository.
Since its release, we have seen contributions to our codebase from members of
the GCC development community and downloads from interested developers.

\section{Evaluation}
\label{sec:eval}
We evaluated last writer slices and \ctraps along several axes.  We show that
last writer slices provide information that is useful during debugging in
Section~\ref{sec:eval:debugging} looking at a set of buggy programs and
comparing to prior, related work.  We evaluate our performance overheads and
show that last writer slices and \ctraps are feasible for use in deployed
systems for a majority of workloads we studied.  We then characterize our
instrumentation and discuss the sources of our system's run time overhead.   We
show that \ctraps is useful by implementing two existing applications and
analyzing their performance.

@@

We illustrate the debugging benefits of last writer slices by using our system
to debug five real-world programs studied previously in the debugging
literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
Section~\ref{sec:eval:debugging:cases} looks in detail at several cases to 
better illustrate how last writer slices helps with debugging.

In Table~\ref{tab:bugs}, we show the bugs we studied describing the program,
bug type, and bug report identifier where applicable.  For each bug, we
followed the procedure for installation and bug triggering described by the
author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
in this work, including both single- and multi-variable bugs and both atomicity
and ordering violations.

@@

\begin{table}
\scriptsize
\centering
\begin{tabular}{l|ll|l|ccc|c}
{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    & Yes   & Yes \\
httrack& 3.43.9  &  N/A     & Order  &Yes  &No    & Yes   & No  \\%Used w/o origin - no help
jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & Maybe\\%All values usable
%apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
\end{tabular}
\caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
\normalsize
\end{table}

The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
violation bugs, last writer slices are effective.  In all cases, we found that
the last writer slice revealed the provenance of the data involved in the
operation that caused the failure, so last writer slicing was effective.  We
found that none of the failures we looked at were trivially debuggable.  This
finding is consistent with the difficulty of debugging concurrency bugs -- the
core dump does not reveal the interaction between threads that led to a
failure, only the result of the interaction, which leads to the failure.

Last writer slices were obviously useful in all five cases.  In these cases,
the root cause of the bug becomes clear when last writer slices show the
inter-thread data flow led to a variable having a value that caused a failure.
Using that information, we expect it would be straightforward in these cases to
add synchronization to fix these bugs.  Note that without last writer slices,
understanding the responsible inter-thread data flow and fixing the bug would
be difficult.

%For the remaining case, {\tt apache2}, last writer slices would
%probably help with debugging, but it is less obvious because this bug is a
%complex, multi-variable bug.  We provide more detail on {\tt apache2} in
%Section~\ref{sec:eval:debugging:cases}.

Our evaluation shows that bad value origin tracking~\cite{badapples} is not
helpful in three out of five cases where last writer slices is
useful.  In one case, a use-after-delete bug in {\tt pbzip2}, bad value origin
tracking helps.  For this bug, when one thread deletes a variable, it makes
that variable unusable and bad value origin tracking would track that value.
When the failure occurs, the programmer would then know why the value was
deleted.

In the remaining case, {\tt apache}, bad value origin tracking may help.  The
bug is an atomicity violation that leads to a double free.  Two threads
manipulate an object's reference count and deallocate the object when the count
hits zero. Zero-checks and deallocation should be atomic, but the code does not
ensure they are.  If threads interleave their code, multiple threads might see
zero reference counts, which is incorrect.  When that happens, both threads try
to free the object and the second free causes a crash.  Bad value origin
tracking may help in this case.  At the first free, the freed pointer becomes
unusable and its origin is tracked.  At the crash, the bad value's reported
origin is the code point that frees the object.  That information potentially
helps debug the program by suggesting a double free.  However, in {\tt apache},
there are many threads, some perform different tasks, and all may free that
object at that code point.  Bad value origin tracking does not record the
thread of the origin.  The programmer is left unsure of whether the problem is
with single-threaded logic or with synchronization.  In contrast, last writer
slices reports the freed value's origin's thread and code point, making it
clear that the programmer is dealing with a concurrency bug, and which threads
are involved.

@@

\paragraph{JDK-StringBuffer}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:jdklws},
and the text in Section~\ref{sec:debugging} describes the benefit provided by
last writer slices in debugging an ordering violation in the JDK-1.4.

@@

\paragraph{HTTrack}
Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of a concurrency bug from HTTrack-3.43.9.  The program crashes at point
{\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid mutex.    The
root cause of this bug is that Thread 1 fails to initialize the mutex before
Thread 2 uses it.   The core dump is of little debugging use, only showing
that, indeed, the lock is uninitialized.  Even if the programmer compared the
core dump from the failing run to the memory image from a correct execution --
a helpful debugging strategy -- they would find only that in a correct
execution, the lock contains valid mutex state.

@@

Looking at last writer slices make this root cause clear: in the failing
execution, the last writer slice indicates the mutex is uninitialized; in the
non-failing execution, the last writer slice indicates the mutex was
initialized by Thread 1 at point {\bf A}. While the core dump provides no
information about the connection between point {\bf A} and point {\bf B}, the
last writer slice reveals that the programmer should ensure {\bf B} follows
{\bf A} to prevent the failure.  Note that even augmenting the core dump with
bad value origin tracking~\cite{badapples} does not help with this bug.  In the
failing execution, origin tracking shows that the unusable lock is
uninitialized, revealing no more than the core dump.  In the non-failing
execution, origin tracking does not monitor the {\em usable} lock value, adding
nothing.  By contrast, comparing the last writer slices for the failing and
non-failing executions reveals how to fix this bug.

@@

We triggered the bug by running an insert query during a log rotation and
execution stopped at {\bf C}.  The core dump showed the log was marked closed,
which is the failure state ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
However, the core dump did not reveal {\em why} the log was marked closed.  The
last writer slice at {\bf C} for {\tt log\_type} shows that it was written last
by Thread 1 at {\bf A}, in the log rotation code, which should be atomic.
Thus, the last writer slice shows the programmer the need for log rotation to
be atomic with respect to the insert code to prevent this failure.  Bad value
origin tracking does not help debug this bug because {\tt log\_type} only ever
contains usable values so it would not be tracked.
